---
title: "Introduction to R and Tidyverse"
output: github_document
---


# Why this workshop?

Fill in the gaps left by traditional econometrics and methods classes using less flexible econometrics and statistical packages.

- Practical skills that tools that will benefit your dissertation (master or doctoral) and future professional career.
- Neglected skills like how to actually find datasets in the wild and clean them.

- Data science skills are largely distinct from (and complementary to) the core 'metrics oeuvre familiar to economists.

- Data viz, cleaning and wrangling; programming; cloud computation; relational databases; machine learning basics

# Some research project examples


- Globalization and suicide: (See the full code and data [here](https://www.kaggle.com/tony75/globalization-and-suicide))
- Predicting the Body Mass Index (BMI). (See the full code and data [here](https://www.kaggle.com/tony75/predicting-bmi))
- Missing values and Clustering: Knowledge Economy in African countries. This is beyond the scope of this course, but it might relevant for data cleaning, and description analysis: (See the full code and data [here](https://github.com/antonio1970/Clustering-Algorithms))
- Gender wage discrimination in China. This is a very simple data visualization of wage distribution across gender: See the full code and data [here](https://www.kaggle.com/tony75/gender-wages)

# Lecture outline

## Introduction to R basics

- Introduction: Motivation, software installation, and data visualization
- Learning to love the shell
- R language basics
- Data cleaning, and wrangling with the "Tidyverse"


## Introduction to machine learning algorithms

- Supervised vs unsupervised learning

- The caret package and regression models

- Clustering techniques

- Complete example: Using African data from the WB and WGI.

# Why R and RStudio? (cont.)

```{R, indeeddotcom, echo = F, fig.height = 6, fig.width = 9, dev = "svg"}
library(tidyverse)
# The popularity data
pop_df <- 
  data.frame(
  lang = c("SQL", "Python", "R", "SAS", "Matlab", "SPSS", "Stata"),
  n_jobs = c(107130, 66976, 48772, 25644, 11464, 3717, 1624),
  free = c(T, T, T, F, F, F, F)
  )
## Plot it
pop_df %>% 
  mutate(lang = lang %>% factor(ordered = T)) %>%
  ggplot(aes(x = lang, y = n_jobs, fill = free)) +
  geom_col() +
  geom_hline(yintercept = 0) +
  aes(x = reorder(lang, -n_jobs), fill = reorder(free, -free)) +
  xlab("Statistical language") +
  scale_y_continuous(label = scales::comma) +
  ylab("Number of jobs") +
  labs(
    title = "Comparing statistical languages",
    subtitle = "Number of job postings on Indeed.com, 2019/01/06"
    ) +
  scale_fill_manual(
    "Free?",
    labels = c("True", "False"),
    values = c("#f92672", "darkslategray")
    ) +
  ggthemes::theme_pander(base_size = 17) +
  # theme_ipsum() +
  theme(legend.position = "bottom")
```



### Data science positivism

- Alongside Python, R has become the *de facto* language for data science.
  - See: [*The Impressive Growth of R*](https://stackoverflow.blog/2017/10/10/impressive-growth-r/), [*The Popularity of Data Science Software*](http://r4stats.com/articles/popularity/)
  
- Open source (free!) with a global user-base spanning academia and industry.

### Bridge to applied economics and other tools

- Already has all of the statistics and econometrics support, and is amazingly adaptable as language to other
programming languages and APIs.

- The RStudio IDE and ecosystem allow for further, seemless integration.

- It's also the language that I know best.



# R vs Stata

If you're coming from Stata, some additional things worth emphasizing:

- Multiple objects (e.g. data frames) can exist happily in the same workspace. 
   - No more `keep`, `preserve`, `restore` hackery.
   - This is a direct consequence of the OOP approach.

- You will load packages at the start of every new R session. Make peace with this.
  - "Base" R comes with tons of useful in-built functions. It also provides all the tools necessary for you to write your own functions. 
  - However, many of R's best data science functions and tools come from external packages written by other users.

- R easily and infinitely parallelizes. For free.
  - Compare the cost of a [Stata/MP](https://www.stata.com/statamp/) license, nevermind the fact that you effectively pay per core...

- You don't need to `tset or xtset` your data. (Although you can too.)



# Examples using ggplot2

```{r gapm_plot, dev="svg"}
library(ggplot2)
library(gapminder) ## For the gapminder data
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
```

# Brief aside: The gapminder dataset

Because we're going to be plotting the [gapminder](https://github.com/jennybc/gapminder) dataset, it is helpful to know that it contains panel data on life expectancy, population size, and GDP per capita for 142 countries since the 1950s.

```{r gapm}
gapminder
```


---

# What else? (cont.)

Elaborate extension: Animation! (See the next slide for the resulting GIF.)
```{r gganim1, eval= TRUE}
library(gganimate)
library(gapminder)
library(gifski)
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```


# What else? (cont.)

There's a lot more to say, but I think we'll stop now for today's lecture.

Rest assured, you will be using ggplot2 throughout the rest of this course and developing your skills along the way.

In the meantime, I want you to do some reading and practice on your own. Pick either of the following (or choose among the litany of online resources) and work through their examples:
- [Chapter 1](https://www.jaredlander.com/r-for-everyone/table-of-contents/) of *R for Everyone: Advance and Analytics and Graphics* by Jared P. Lander.
- [Chapter 1](https://r4ds.had.co.nz/introduction.html) of *R for Data Science* by G. Grolemund and H. Wickham. (Or just try a chapter at random.) 



# First steps

1. Open RStudio on yours or a group member's laptop
2. Check RStudio version (version command, mine )
2. Click File > New File > R script. 

You will see a blank section of screen in the top-left of your RStudio window. This is where you will write your first R script. This is similar to write a do file in Stata. Notice that R code is saved as a `.r file`.

# Four basic areas on your screen

1. The is the `script` editor as in Stata
2. The bottom left is the `console` (you can use R interactively by typing commands as in Stata)
3. The upper right is where you see the list of objects in your workspace. History is also there (you know at aytime what you did type in your console)
4. The bottom right is where you find the available packages, graphical output, your file structure, and help.  

### Help

Getting help with R is easy. To get further information about the syntax of any R function,  just type a `?` before the function name. See the example below

```{r}
# regression function (fitting linear models by OLS)
?lm
```


If you don't know exactly the name of the fuction, just use double question marks

```{r}
??cluster
help.search("cluster")
apropos("cluster")
```


Finally, the structure of the help for each R function is:

1. Description
2. Usage
3. Arguments
4. Value
5. References

### Console
The bottom-left of the screen has a tab called "Console". This is basically a very fancy calculator.

Try the basic calculator by typing something like
```{r}
2+2
```

Or even something fancier like
```{r}
sqrt(pi)
b <- 1.4156
sqrt(b)
```


### More options


```{r decimals}
# We can use the command options digits to set up the number of decimals as you wish
options(digits = 3)
sqrt(10)
ceiling(3.5) # round up to next integer
floor(3.5) # round down to the next integer
log(100)  # natural log
exp(2)  # exponents
```

You can also test for equality. True and false are the keywords. Notice that T and F are synonyms but are discouraged.
TRUE 
FALSE

```{r}
1==2 # Equality return TRUE if equal
1 !=2 # Return false if unequal
10 >= 9  # Greater than
4<=1 # Less than
```


### Objects and basic R

R is a oriented object language. For R, everything is an object, and each object has its own name.

1. To remove any particular object from the current session employ `rm()`
2. To list all objects in your workspace use `ls()`
3. To remove all objects from your workingspace just type `rm(list =ls())`

```{r}
x<-2 # Creating an object
ls()
rm(list = ls())
```


Lastly, most of the them when using variables in a dataframe, you refer to them by using the `$`symbol. This can be avoided by using the `attach` command. This will be explained more in detail in next labs.

Just as an example, let us load one of the built in datasets in R, `mpg´ from the **ggplot2 package**. This dataset contains observations collected by the US Environmental Protection Agency on 38 models of cars. 
```{r}
data(mtcars)
mean(mtcars$mpg) # Notice that we get access to the variables with the dollar symbol
attach(mtcars)
mean(mpg) # there is no need to use the dollar symbol
```

### Working directory

```{r}
getwd() # to check your current working directory
# setwd() # You can also use the menu  session and then set working directory, otherwise you need to specify the entire path
```

### Packages
R makes extensive use of third-party packages. We will not get into the details right now, but for this class, you will need to install a few of these. Installing packages is quite easy. Type the following two lines of code at the very top of your script:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
install.packages("tidyverse", repos='http://cran.us.r-project.org')
install.packages("skimr", repos='http://cran.us.r-project.org')
install.packages("wooldridge", repos='http://cran.us.r-project.org')
install.packages("learnr", repos='http://cran.us.r-project.org')
```

You've just installed two basic packages. We can also check it out by using the following syntax

```{r, eval= FALSE}
installed.packages()
```




Basically, you have downloaded them onto your laptop. Just like with other software on your computer, you only need to do the installation once. However, you still need to tell R that you will be using the packages (by using the `library`command). Add the following two lines of code to your script (below the first two lines you wrote). Notice how there are no quotation marks inside the parenthese this time.

Whenever you start over a new Rstudio session, you need to load the appropriate libraries.

In RStudio, you can also install packages by using the R interface. You will find it at Tools -> Install Package, and there you will get a pop--up window to type the package you want to install


### Running a script
To execute the script, click on the word "Source" in the top-right corner of the top-left window pane. This will take what is in your script and automatically send it to the console (as if you typed it directly into the console)

To save the script, click on the disk icon at the top of your script pane (but not the disk icon at the very top of RStudio). Name your script `ICL1_XYZ.R` where `XYZ` are your initials.

### Commenting
Now, put a hashtag (#) in front of the first two lines of code in your script, like so:
```{r}
#install.packages("tidyverse")
#install.packages("skimr")
#install.packages("wooldridge")
```

The hashtag is how you tell R not to run the code in your script. This is known as "commenting" your code.

At the very top of your script, type the names of your group members with a hashtag in front.

### Citation R and packages 

If you need to cite R, just use

```{r}
citation()
```

if for instance, you need to cite one of the packages, then use

```{r}
citation(package = "ggplot2")
```